---
title: 'GENE360: 1000 Genomes Project Module'
author: "Associate Professor Mik Black"
date: "18 April 2018"
output: html_document
---

## Overview

The objective of this module is to give you some familiarity with the types of data and tools that are commonly used for genetic studies of population diversity.

The data we will be using are publicly available high-throughput sequencing 
data from the 1000 Genomes Project.  The main web page for the 1000 Genomes Project is:

http://www.internationalgenome.org/

The following page provides background information about the 1000 Genomes Project (and is worth reading, particularly the information about the populations being used in the study):

http://www.internationalgenome.org/about/

We will be using the R and RStudio software applications to analyze the data. 

R: https://cran.rstudio.com/ <BR>
RStudio: https://www.rstudio.com/product/rstudio/download/

### Getting started - loading the data

To get started, we need to load some data.  I have placed a zip file containing
the data files on the GENE360 Blackboard page.  Download and unzip this file - then we need to tell RStudio where to find the data.  This involves "setting the working directory" for R - choose "Set Working Directory" and then "Choose Directory" from the "Session" menu in RStudio, and then find the folder containing the module data and click "Open".

The first data set we will look at contains genetic data on just a few loci (specific points in the genome) across the various populations being studied.  Looking at this small data set will give you a feel for what we mean by "genotype data" for a particular locus.

Open RStudio and read in the first data set using the following command:

```{r}
## Read in the SNP data
snpData = read.table('GENE360snpData.txt',sep='\t',header=T)
```

The ## characters indicate a comment - anything typed after them is ignored by R.  The next line tells R to read in data from the file "GENE360snpData.txt", which it calls "snpData". 

In this handout, the "code" used to run commands in R (and any output produced) is indicated by `courier font`.  The "output" from each command appears below the code, prefixed by the ## symbols.

We can ask R about the characteristics of the snpData object.  The command:

```{r}
dim(snpData)
```

will tell us how many rows (individuals) and columns the data set contains.  The output shows us that there are 2504 rows and 11 columns.

\pagebreak 
The column names can be found using the following command:

```{r}
names(snpData)
```

To look at the full data set, you can use the "View" command:

```{r, eval=FALSE}
## Can view the data file within RStudio via 
View(snpData)
```

The second column is called "Population".  We can make a table of this information to see how many individuals are present in each population (the "$" sign tells R to use the "Population" column from the "snpData"" object):

```{r}
table(snpData$Population)
```

The super-population codes are:

AFR, African; AMR, Ad-Mixed American; EAS, East Asian; EUR, European; SAS, South Asian.

Additional information about the composition of these populations can be found at:

http://www.internationalgenome.org/faq/which-populations-are-part-your-study

## Looking at SNP frequencies

The other nine columns in the data set relate to specific single nucleotide polymorphisms (SNPs) in the genome - we have the genotype data for each SNP for every individual in the data set.

We can use the "table" command again to summarize the genotype information for each SNP:

```{r}
## Make a genotype frequency table for the first SNP
table(snpData$rs3826656)
```

We can also calculate the proportions associated with each genotype:

```{r}
## Calculate proportions
prop.table(table(snpData$rs3826656))
```

and examine differences in genotype frequencies across populations:

```{r}
## Create contingeny table - genotypes across populations
table(snpData$Population, snpData$rs3826656)

## And calculate proportions (rounded)
round(prop.table(table(snpData$Population, snpData$rs3826656),1),2)
```

These results can be plotted as a bar plot:

```{r}
snpFreqs = t(prop.table(table(snpData$Population, snpData$rs3826656), 1))
barplot(snpFreqs, beside=TRUE, legend.text=TRUE, ylim=c(0,1))
```

These analyses can be repeated for different SNPs by changing the SNP ID (e.g., rs3826656) in the above commands.

### Why look at these SNPs?

The nine SNPs contained in this data set were not just randomly chosen - they are SNPs that have been found to be associated with altered disease risk.  That is, an individual's genotype at a particular position in the genome affects their risk of developing a particular disease.  This is not absolute though, it really just raises or lowers the probability of disease - it doesn't guarantee complete protection or susceptibility.

SNP|Alleles|Nearby Gene|Disease
---|-------|----|-------|----
[rs3826656](http://www.snpedia.com/index.php/Rs3826656)|A/G|CD33|Alzheimer's Disease
[rs13387042](http://www.snpedia.com/index.php/Rs13387042)|A/G|DIRC3|Breast Cancer
[rs4779584](http://www.snpedia.com/index.php/Rs4779584)|T/C|GREM1|Colorectal Cancer
[rs2398162](http://www.snpedia.com/index.php/Rs2398162)|A/G|NR2F2|Hypertension
[rs7299940](http://www.snpedia.com/index.php/Rs7299940)|C/G| - |Panic Disorder
[rs1344706](http://www.snpedia.com/index.php/Rs1344706)|A/C|ZNF408A|Schizophrenia and Bipolar Disorder
[rs7659604](http://www.snpedia.com/index.php/Rs7659604)|T/C|TMEM155|Type 2 Diabetes
[rs734553](http://www.snpedia.com/index.php/Rs734553)|T/G|SLC2A9|Gout
[rs113010081](http://www.snpedia.com/index.php/Rs113010081)|C/T|CCR5|HIV/AIDs "resistance"

For each SNP, increased risk of disease (or for rs113010081, decreased risk) is associated with the 
minor allele.  Variation in genotype frequencies across populations can help to explain some of the population-specific differences in rates of different diseases.

Clicking on SNP IDs in the table above will link through to aditional information about each variant.

## Population diversity

The second data set that we will look at can be used to investigate population (genetic) diversity.  These SNPs were chosen because they are "ancestry informative" - variation in their genotype frequencies tends to be associated with differences between population groups.

```{r, cache=TRUE}
## Load ancestry data
snpAns = read.table('GENE360snpAncestry.txt', header=T, sep='\t')
## Check dimensionality of data
dim(snpAns)
```

Now we have a lot more data: 2504 individuals, and 2302 SNP genotypes for each.  Note that there are 2305 columns in the data set, but the first three relate to non-SNP data.  

Have a look using the View command:

```{r, eval=FALSE}
## Can view the data file within RStudio via 
View(snpAns)
```

\pagebreak
Note that we also have data about the sub-populations now:

```{r}
table(snpAns$SubPopulation, snpAns$Population)
```

To examine population diversity, we need to do two things:

1. create a data object of ONLY the SNP genotype data (i.e., remove the first three columns).
2. convert the genotypes to allele counts (e.g., TT, AT, AA to 0, 1, 2 - count the number of A's).

Step 1:
```{r}
## Create object containing only the SNP data - remove the first three columns
snpAnsDat = snpAns[,-c(1,2,3)]
```

Step 2:
```{r, cache=TRUE}
## Load a custom function to convert to allele counts
source('alleleCounts.R')

## Apply the function to the genotype data, one column (SNP) at a time
## This will take 30 seconds or so...
snpAnsCount = apply(snpAnsDat, 2, alleleCounts)
```

Now we have a data set of just the SNP data, with genotypes converted to allele counts.

Have a look at the new data set using the View command:

```{r, eval=F}
View(snpAnsCount)
```

So, what are we going to do with this new data set?

### Dimension reduction (principal components analysis)

In terms of examining population diversity, we have 2302 dimensions of data available - one dimension for each SNP.

Rather than trying to comprehend this huge amount of data in 2302-dimensional space, genetics researchers often use a statistical tool called Principal Components Analysis (PCA) to reduce the dimensionality of the data.

The idea is to find the most important variation in the data, and examine the samples in terms of that variation, ignoring the rest.   In practice, this works fairly well, because genetic differences between populations provide a strong (and relatively consistent) source of variation across genomic loci (i.e., SNPs).  Rather than looking at 2302 dimensions of data, we end up looking at variation across just 2 or 3 dimensions - each dimension is defined by a combination of SNPs which vary in a similar way across the individuals in the study.

```{r, cache=TRUE}
## Load custom function to perform principal components analysis
source("pcaGenotypes.R")

## Apply this function to the allele count data
## This will take a couple of minutes to run
pca = pcaGenotypes(snpAnsCount)
```

To visualise the populations across the principal components, we need to define colours for each populations. My arbitrary choice (slightly influenced by the 
Y-chomosome paper from last week) was: AFR (brown), AMR (red), EAS (purple), EUR (blue), SAS (green).

```{r}
## Create an object relating to the population data
ansPop = snpAns$Population

## Generate colours to associate with each population
pCols = c("brown","red","purple","blue","green")
popCol = pCols[as.numeric(as.factor(ansPop))]
names(popCol) = ansPop

## Check that they correspond to populations
table(popCol,ansPop)
```

Now we can plot the principal components and colour the points based on the population each sample belongs to.

```{r, fig.height=7}
## Load a custom plotting function to generate the plots:
source('plotPCA.R')

## Plot the PCA data that we have generated, along with the population information
plotPCA(pca, popCol)
```

From the plots we can see that samples from the same population tend to cluster together, and that the first three principal components do a reasonable job of capturing the genetic diversity between the populations.

With the `scatterplot3d` package, you can plot the first three principal
components at once (i.e., combining the information from the three
scatterplots above).  This shows that the European (EUR), East Asian (EAS) and 
South Asian (SAS) super-populations are relatively homogeneous, while the Ad-Mixed American (AMR) and African (AFR) super-populations exhibit greater variation, suggesting admixture within these groups.

```{r}
library(scatterplot3d)
scatterplot3d(pca[,1], pca[,2], pca[,3], color=popCol, pch=16,
              cex.symbols=0.5, xlab="PC1", ylab="PC2", zlab="PC3")
```

## Phylogenetic trees

Another way to visualise genetic similarity is via "phylogenetic trees", also known more generally as "dendrograms".

These tree diagrams group items together based on similarity scores.  In this setting our "items" are the populations, and the scores are calculated based on the genetic similarities between each pair of populations.

To calculate similarities, we need to create a set of "average" genotypes for each population.  One way to do this is to calculate the frequency of the major homozygote for each SNP in each population.

```{r, cache=TRUE}
## Load custom function for calculating major homozygote frequency:
source('calcMajorFreq.R')

## Apply this function to the SNP count data
## Takes a few seconds
popFreqs = calcMajorFreq(snpAnsCount, ansPop)
```

```{r, fig.height=4}
## Make a cluster tree based on these frequencies
plot(hclust(dist(popFreqs)),hang=-1)
```

The way in which the "leaves" of the tree cluster, reflects the similarity between the items analysed.  In this case the populations are grouped based on their genetic similarity, as measured by these particular loci.  Here it appears that the AMR (Ad-Mixed American), and SAS (South Asian) super-populations are most genetically similar, then EUR (European), followed by EAS (East Asian).  The AFR (African) super-population is the most genetically dissimilar relative to the others.  It is likely that these groupings reflect (to some degree) migration patterns and shared ancestry experienced by these populations.

\pagebreak

## Assessment tasks

Use the data and code from above to answer the following questions.  Hand in your code and the output generated, along with your answers to the questions below.  

1. Select one of the nine SNPs from the first data set and investigate its variation (in terms of genotype frequencies) across the populations of the 1000 Genomes Project.  Briefly overview the support for this SNP being a modifier of
disease risk, and try to find evidence (in the literature) for variation in population-specific risk as a result of differences in genotype frequency.

2. Create a plot of PC1 _versus_ PC2, colouring points by sub-population rather than super-population.  Rather than use my `plotPCA` function, you will need to use R's generic plotting commands.  The following code generates a PCA plot from scratch, using the super-population colours:
```{r, eval=FALSE}
plot(pca[,1], pca[,2], col=popCol, pch=16, cex=0.5, xlim=1.5*range(pca[,1]))
legend("topright", legend=names(table(snpAns$Population)), fill=pCols, cex=0.5)

## Notes on parameters:
##  - "col" specifies the colors
##  - "pch" specifies the shape of the plotting symbol
##  - "cex" specifies the size of the points (or text, for the legend)
##  - "xlim" specifies the range of the x axis
##  - "fill" specifies the colours to be used in the legend
```
Loading the `subPopColours.RData` file via:
```{r, eval=FALSE}
load('subPopColours.RData')
```
will provide two colour objects: `sbCols` and `subPopCol`, which you should be able to use to generate a PCA plot with the points coloured by sub-population.  

Comment on the distribution of the sub-populations with respect to genetic variability.  Which of the sub-populations exhibit evidence of admixture, and which populations (sub- or super-) are the most likely contributors?  Is there support for these findings in the literature?  Provide a brief summary of any relevant publications you find.

Generate similar plots of PC1 _versus_ PC3 and PC2 _versus_ PC3.  Based on these plots, describe the main population features that are captured by each of the first three principal components.

3. Generate a phylogentic tree for the sub-population data.  Comment on how well the structure of the tree agrees with the relationships between the super- and sub-populations of the 1000 Genomes Project data.  Are there any super-populations that are split across different parts of the tree?  If so, what is the likely cause?  Discuss the shape of the tree (and the placement of the super- and sub-populations) in the context of the "out of Africa" theory of human migration. How does the shape of the tree support (or question) the proposed migration path?  Also comment of the relationship between the shape of the tree, and the distribution of the populations in the PCA plot you generated in (2).  What advantages and disadvantages do you feel the two plotting methods (PCA plot and phylogenetic tree) have when visualising genetic diversity data?

